---
title: "What's worth keeping?" 
subtitle: "An analysis of Danish culture heritage management practices over time"
author: "Laura Emilie Hollænder Jensen"
date: "2026-05-25"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### Loading packages and data
```{r}
library(tidygeocoder)
library(tidyverse)
library(ggplot2)
library(leaflet)
library(leaflet.extras)
library(htmltools)
library(htmlwidgets)
library(mapview)
#webshot::install_phantomjs()

buildings <- read_csv2("data/listed_buildings_geocoded_english.csv")
glimpse(buildings)
```

# 1. Mapping listed buildings in Denmark
### Map 1: With clusters to show geographical dispersion of listed buildings.
```{r}
# Removing NAs for mapping 
buildings_for_mapping <- buildings %>% 
  filter(!is.na(Latitude) & !is.na(Longitude))

# Making a leaflet map
cluster_map <- leaflet(buildings_for_mapping) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
    addCircleMarkers(
    lng = ~Longitude, 
    lat = ~Latitude, 
    radius = 5,
    fillColor = "#2B5C8F", 
    fillOpacity = 0.8,
    stroke = TRUE, 
    weight = 1, 
    color = "#FFFFFF",     
        popup = ~paste0(
      "<h3>", htmlEscape(Building), "</h3>",
      "<b>Address:</b> ", htmlEscape(Address), "<br>",
      "<b>Municipality:</b> ", htmlEscape(Municipality), "<br>",
      "<b>Year of Construction:</b> ", htmlEscape(ifelse(is.na(Year_of_construction) | Year_of_construction == "", "Unknown", Year_of_construction)), "<br>",
      "<b>Year of Listing:</b> ", htmlEscape(ifelse(is.na(Year_of_listing) | Year_of_listing == "", "Unknown", Year_of_listing)), "<br>",
      "<b>Architectural Period:</b> ", htmlEscape(ifelse(is.na(Period) | Period == "", "Unknown", Period))
    ),
        clusterOptions = markerClusterOptions(
      disableClusteringAtZoom = 16, 
      maxClusterRadius = 40         
    )
  )

# Take a snapshot of the map
mapshot(
  cluster_map, 
  file = "cluster_map_snapshot.png",
  vwidth = 1600,    
  vheight = 1200,   
  zoom = 2           
)

cluster_map
```

### Map 2: With toggle option to visualise the spread of listed buildings by architectural period.
```{r}
buildings_for_mapping <- buildings_for_mapping %>%
  mutate(
    Period = ifelse(is.na(Period), "Year of construction unknown", as.character(Period)),
    Period = factor(Period, levels = c(
      "Medieval", "Renaissance", "Baroque and rococo", 
      "Classicism", "Historicism", "Modernism", "Year of construction unknown"
    ))
  )

period_colors <- colorFactor(
  palette = c(
    "Medieval"                     = "#2D1610", 
    "Renaissance"                  = "#8B1E14", 
    "Baroque and rococo"           = "#C25E1A", 
    "Classicism"                   = "#D9A74A", 
    "Historicism"                  = "#2B5C8F", 
    "Modernism"                    = "#2E8B57", 
    "Year of construction unknown" = "#71717A"  
  ),
  domain = buildings_for_mapping$Period,
  ordered = TRUE
)

# Making a leaflet map
toggle_map <- leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron)

# Adding every period as an independent layer
period_list <- levels(buildings_for_mapping$Period)

for (p in period_list) {
  data_sub <- buildings_for_mapping %>% filter(Period == p)
  
  if (nrow(data_sub) > 0) {
    toggle_map <- toggle_map %>%
      addCircleMarkers(
        data = data_sub, 
        lng = ~Longitude, lat = ~Latitude, radius = 5,
        fillColor = ~period_colors(Period), fillOpacity = 0.8,
        stroke = TRUE, weight = 1, color = "#FFFFFF",
        popup = ~paste0(
          "<h3>", htmlEscape(Building), "</h3>",
          "<b>Period:</b> ", htmlEscape(Period), "<br>",
          "<b>Address:</b> ", htmlEscape(Address), ", ", htmlEscape(Municipality), "<br>",
          "<b>Year of Construction:</b> ", htmlEscape(ifelse(is.na(Year_of_construction) | Year_of_construction == "", "Unknown", Year_of_construction)), "<br>",
          "<b>Year of Listing:</b> ", htmlEscape(ifelse(is.na(Year_of_listing) | Year_of_listing == "", "Unknown", Year_of_listing))
        ),
        group = p 
      )
  }
}

# Adding toggle switch box for the different architectural periods.
toggle_map <- toggle_map %>%
  addLayersControl(
    overlayGroups = period_list,
    options = layersControlOptions(collapsed = FALSE),
    position = "topright"
  ) %>%
  addLegend(
    pal = period_colors, 
    # FIX: Updated data source reference
    values = buildings_for_mapping$Period, 
    title = "Architectural period",
    position = "bottomright"
  ) 

# Take a snapshot of the map
mapshot(
  toggle_map, 
  file = "toggle_map_snapshot.png",
  vwidth = 1600,     
  vheight = 1200,    
  zoom = 2          
)

# Display map
toggle_map
```

# 2. Plotting listed buildings in Denmark
### Plot 1: Listed buildings in Denmark by architectural period
```{r}
buildings_for_plotting <- buildings %>%
  filter(!is.na(Period)) %>% 
  count(Period) %>% 
    mutate(Period = factor(Period, levels = c(
    "Medieval", "Renaissance", "Baroque and rococo", 
    "Classicism", "Historicism", "Modernism"
  ))) %>% 
    arrange(Period)

plot_periods <- ggplot(buildings_for_plotting, aes(x = Period, y = n, fill = Period)) +
  geom_col(show.legend = FALSE, width = 0.65) +
  geom_text(aes(label = n), vjust = -0.5, size = 4, family = "Georgia") +
    scale_x_discrete(labels = c(
    "Baroque and rococo" = "Baroque & rococo\n(1650-1749)",
    "Classicism"         = "Classicism\n(1750-1849)",
    "Historicism"        = "Historicism\n(1850-1944)",
    "Medieval"           = "Medieval\n(-1536)",
    "Modernism"          = "Modernism\n(1945-)",
    "Renaissance"        = "Renaissance\n(1536-1649)"
  )) +
    scale_fill_manual(values = c(
    "Medieval"                     = "#2D1610", 
    "Renaissance"                  = "#8B1E14", 
    "Baroque and rococo"           = "#C25E1A", 
    "Classicism"                   = "#D9A74A", 
    "Historicism"                  = "#2B5C8F", 
    "Modernism"                    = "#2E8B57"
  )) +
    labs(
    title = "Listed buildings in Denmark by architectural period",
    x = "Architectural period",
    y = "Number of buildings",
    caption = "Data: Slots- og Kulturstyrelsen (2026)"
  ) +
  
  theme_minimal() +
  theme(
    text = element_text(family = "Georgia"),
    plot.title = element_text(face = "bold")
  )

# Display and save
print(plot_periods)
ggsave(filename = "architectural_periods_plot.png", plot = plot_periods, width = 8, height = 5)

```

### Plot 2: Listings over time
```{r}
plot_listings <- buildings %>% 
  filter(!is.na(Year_of_listing),
         Year_of_listing >= 1918, 
         Year_of_listing <= 2026) %>%
  ggplot(aes(x = Year_of_listing)) +
  geom_histogram(
    binwidth = 5,
    fill = "#8B1E14", color = "white", linewidth = 0.4
  ) +
  scale_x_continuous(breaks = seq(1900, 2026, 10)) +
  scale_y_continuous(breaks = seq(0, 1000, 100)) +
  labs(
    title    = "When were buildings granted protected status?",
    subtitle = "Each bar represents five years",
    x = "Year of listing", y = "Number of buildings"
  ) +
  theme_minimal() +
  theme(
    text = element_text(family = "Georgia"),
    plot.title = element_text(face = "bold"))
  
print(plot_listings)
ggsave(filename = "listings_over_time_plot.png", plot = plot_listings, width = 8, height = 5)

```

### Plot 3: Building age at listing over time
```{r}
five_year_summary <- buildings %>% 
  mutate(
    Year_of_listing = as.numeric(Year_of_listing),
    Year_of_construction = as.numeric(Year_of_construction)
  ) %>% 
  filter(!is.na(Year_of_listing), !is.na(Year_of_construction),
         Year_of_listing >= 1918, Year_of_listing <= 2026,
         Year_of_construction > 0) %>% 
  
  mutate(Age_at_listing = Year_of_listing - Year_of_construction) %>% 
  filter(Age_at_listing >= 0) %>% 
  
  mutate(Five_Year_Period = floor(Year_of_listing / 5) * 5) %>% 
  
  group_by(Five_Year_Period) %>% 
  summarise(
    Median_Age = median(Age_at_listing),
    Total_Listings = n()
  )

plot_five_year_trend <- ggplot(five_year_summary, aes(x = Five_Year_Period, y = Median_Age)) +
  
  geom_line(color = "darkred", size = 1) + 
  geom_point(aes(size = Total_Listings), color = "darkred", alpha = 0.8) +
  
  # Space out x-axis markings every 10 years so the labels look clean
  scale_x_continuous(breaks = seq(1910, 2026, 10)) +
  
  labs(
    title = "How old were buildings when they were listed?",
    subtitle = "Development in building age at year of listing",
    x = "Year of listing",
    y = "Median age at listing (years)",
    size = "Buildings\nlisted",
    caption = "Data: Slots- og Kulturstyrelsen (2026)"
  ) +
  
  theme_minimal() +
  theme(
    text = element_text(family = "Georgia"),
    plot.title = element_text(face = "bold")
  )

# Print and save
print(plot_five_year_trend)
ggsave("listings_5_year_bubbles.png", plot = plot_five_year_trend, width = 8, height = 5)
```

### Plot 4: Top 10 municipalities by listing numbers over time
```{r}
library(tidyverse)
library(tidytext)

academic_font <- "Georgia"

#Calculate the top 10 municipalities for each decade
top_municipalities_by_decade <- buildings %>% 
  mutate(
    Year_of_listing = as.numeric(Year_of_listing)
  ) %>% 
  filter(!is.na(Year_of_listing), !is.na(Municipality)) %>% 
  mutate(Decade = floor(Year_of_listing / 10) * 10) %>% 
  filter(Decade %in% c(1910, 1920, 1930, 1940, 1950, 1960, 1970, 1980, 1990, 2000, 2010, 2020)) %>% 
  
  # Count listings per municipality per decade
  count(Decade, Municipality) %>% 
  
  # Use the top 10 for each decade
  group_by(Decade) %>% 
  slice_max(n, n = 10, with_ties = FALSE) %>% 
  ungroup() %>% 
  mutate(Decade_Label = ifelse(Decade == 2020, "2020–2026", paste0(Decade, "–", Decade + 9))) %>%
  
# Reorder the municipalities by number of listings in descending order
mutate(Municipality = reorder_within(Municipality, n, Decade))

# Put together the plot using facet_wrap
plot_decade_leaderboard <- ggplot(top_municipalities_by_decade, aes(x = n, y = Municipality)) +
  geom_col(fill = "darkred", color = "white", width = 0.7) +
  geom_text(aes(label = n), hjust = -0.2, size = 2.5, family = "Georgia") +
  facet_wrap(~ Decade_Label, scales = "free_y", ncol = 4) +
  scale_y_reordered() +
  scale_x_continuous(expand = expansion(mult = c(0, 0.3))) +
  labs(
    title = "Which municipalities dominated heritage listings?",
    subtitle = "Top 10 municipalities across decades",
    x = "Number of buildings listed",
    y = NULL,
    caption = "Data: Slots- og Kulturstyrelsen (2026)"
  ) +
  theme_minimal() +
  theme(
    text = element_text(family = "Georgia"),
    plot.title = element_text(face = "bold", size = 14),
    panel.grid.major.y = element_blank(), # Drops unnecessary horizontal line clutter
    strip.text = element_text(face = "bold", size = 8.5), # Bolds the decade panel labels
    panel.spacing = unit(1, "lines"),
    axis.text.x = element_text(size = 6)
  )

# Print and save
print(plot_decade_leaderboard)
ggsave("municipality_leaders_by_decade.png", plot = plot_decade_leaderboard, width = 8, height = 6)
```

