---
title: "Danish Kings"
author: "Laura Emilie Hollænder Jensen"
date: "2026-03-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


The task here is to load your Danish Monarchs csv into R using the `tidyverse` toolkit, calculate and explore the kings' duration of reign with pipes `%>%` in `dplyr` and plot it over time.

## Load the kings

Make sure to first create an  `.Rproj` workspace with a `data/` folder where you place either your own dataset or the provided `kings.csv` dataset. 

1. Look at the dataset that are you loading and check what its columns are separated by? (hint: open it in plain text editor to see)
List what is the separator. 
ANSWER: In Danish_kings.csv, the separator is ";".

2. Create a `kings` object in R with the different functions below and inspect the different outputs.
-  `read.csv()`
-  `read_csv()`
-  `read.csv2()` 
-  `read_csv2()`
ANSWER: In the code chunks below, I will load a kings-object using four different functions. 
I will use the function 'glimpse()' to inspect the output.

```{r eval=TRUE}
library(tidyverse)
library(dplyr)
# FILL IN THE CODE BELOW and review the outputs
kings1 <- read.csv("data/Danish_kings.csv")
glimpse(kings1)
class(kings1)
ncol(kings1)

kings2 <- read_csv("data/Danish_kings.csv")
glimpse(kings2)
class(kings2)
ncol(kings2)

kings3 <- read.csv2("data/Danish_kings.csv")
glimpse(kings3)
class(kings3)
ncol(kings3)

kings4 <- read_csv2( "data/Danish_kings.csv")
glimpse(kings4)
class(kings4)
ncol(kings4)


```
Answer: 
1. Which of these functions is a `tidyverse` function? Read data with it below into a `kings` object

ANSWER: The functions read_csv and read_csv2 are tidyverse functions. 

2. What is the result of running `class()` on the `kings` object created with a tidyverse function.


ANSWER: Running 'class()' on the 'kings' object returns this result: 
"spec_tbl_df" "tbl_df"      "tbl"         "data.frame" 

3. How many columns does the object have when created with these different functions?

ANSWER: The objects created using read.csv and read_csv have only one columns. 
The objects created using read.csv2 and read_csv2 have 5 columns.

4. Show the dataset so that we can see how R interprets each column

ANSWER: The 'glimpse()' function lets me inspect the dataset and shows me how R interprets the data.
The 'Name' column is interpreted as the class character. 
read.csv2 interprets the columns 'Born', 'Dead', 'Reign_start' and 'Reign_end' as the type "integer". 
read_csv2 interprets the columns 'Born', 'Dead', 'Reign_start' and 'Reign_end' as the type "double". 


```{r eval = TRUE}
# COMPLETE THE BLANKS BELOW WITH YOUR CODE, then turn the 'eval' flag in this chunk to TRUE.

# In this code chunk, I am using the function 'read_csv2' to create the object
# that I will perform calculations on. I will call it 'kings'.

kings <- read_csv2("data/Danish_kings.csv")
glimpse(kings)
class(kings)
ncol(kings)
# The function ncol() run on 'kings' returns the number 5. 

```


## Calculate the duration of reign for all the kings in your table
You can calculate the duration of reign in years with `mutate` function by subtracting the equivalents of your `startReign` from `endReign` columns and writing the result to a new column called `duration`. But first you need to check a few things:

* Is your data messy? Fix it before re-importing to R
* Do your start and end of reign columns contain NAs? Choose the right strategy to deal with them: `na.omit()`, `na.rm=TRUE`, `!is.na()`

Create a new column called  `duration` in the kings dataset, utilizing the `mutate()` function from tidyverse. Check with your group to brainstorm the options.

```{r}
# YOUR CODE
kings <- kings %>% 
  mutate(Duration = Reign_end - Reign_start)
glimpse(kings)
```

## Calculate the average duration of reign for all rulers
Do you remember how to calculate an average on a vector object? If not, review the last two lessons and remember that a column is basically a vector. So you need to subset your `kings` dataset to the `duration` column. If you subset it as a vector you can calculate average on it with `mean()` base-R function. If you subset it as a tibble, you can calculate average on it with `summarize()` tidyverse function. Try both ways!

* You first need to know how to select the relevant `duration` column. What are your options?
* Is your selected `duration` column a tibble or a vector? The `mean()` function can only be run on a vector. The `summarize()` function works on a tibble.
* Are you getting an error that there are characters in your column? Coerce your data to numbers with `as.numeric()`.
* Remember to handle NAs: `mean(X, na.rm=TRUE)`


```{r}
# YOUR CODE
# Base R version: 
# First way: Extracting a column as a vector using $. 
mean_duration <- mean(kings$Duration, na.rm = TRUE)
mean_duration

# Tidyverse version:
tidy_mean_duration <- kings %>%
  summarise(mean(Duration, na.rm = TRUE))
tidy_mean_duration

# Returns a tibble containing the mean. 

```

## How many and which kings enjoyed a longer-than-average duration of reign?

You have calculated the average duration above. Use it now to `filter()` the `duration` column in  `kings` dataset. Display the result and also  count the resulting rows with `count()`

ANSWER: 25 kings enjoyed a longer-than-average duration of reign. 

```{r}
# YOUR CODE
kings %>% 
  filter(Duration > tidy_mean_duration) %>% 
  select(Name, Duration)
# This does not work and prints a warning message, because 
# tidy_mean_duration is a tibble i.e. a one column matrix. 
# Let's try it again with the mean I calculated using base R, mean_duration,
# which is a value and not a tibble. 

kings %>% 
  filter(Duration > mean_duration) %>% 
  select(Name, Duration)
# This works! Now, let's count them.

kings %>% 
  filter(Duration > mean_duration) %>% 
  select(Name, Duration) %>% 
  count()

# The 'count()' function returns a tibble containing just one integer: 25. 
# That means that 25 kings enjoyed a longer-than-average duration of reign. 
# We can also make an object containing only the longer-than-average reigning
# monarchs. 

above_average_kings <- kings %>% 
  filter(Duration > mean_duration) %>% 
  select(Name, Duration)
  view(above_average_kings)
  
```


## How many days did the three longest-ruling monarchs rule? 

* Sort kings by reign `duration` in the descending order. Select the three longest-ruling monarchs with the `slice()` function
* Use `mutate()` to create `Days` column where you calculate the total number of days they ruled
* BONUS: consider the transition year (with 366 days) in your calculation!

ANSWER: The three longest-ruling monarchs were Christian IV, Margrethe II and Erik of Pomerania. 
Accounting for leap years, Christian IV ruled for 21915 days, Magrethe II ruled for 18993 days, and Erik of Pomeri

```{r}
# YOUR CODE
# Sorting in the descending order.
kings %>% 
  select(Name, Duration) %>% 
  arrange(desc(Duration))

# Selecting only the three longest-ruling monarchs.
top_kings <- kings %>%
  arrange(desc(Duration)) %>%
  slice(1:3)

# Using mutate() to create a 'Days' column. 
top_kings <- top_kings %>%
  mutate(Days = Duration * 365)


# BONUS: consider the transition year (with 366 days) in your calculation!
top_kings <- top_kings %>%
  mutate(
    Leap_years = Duration %/% 4,
    Days_accounting_for_leap_years = Duration * 365 + Leap_years)

```


## Challenge: Plot the kings' duration of reign through time

What is the long-term trend in the duration of reign among Danish monarchs? How does it relate to the historical violence trends ? 

* Try to plot the duration of reign column in `ggplot` with `geom_point()` and  `geom_smooth()`
* In order to peg the duration (which is between 1-99) somewhere to the x axis with individual centuries, I recommend creating a new column `midyear` by adding to `startYear` the product of `endYear` minus the `startYear` divided by two (`startYear + (endYear-startYear)/2`).
* Now you can plot the kings dataset, plotting `midyear` along the x axis and `duration` along y axis
* BONUS: add a title, nice axis labels to the plot and make the theme B&W and font bigger to make it nice and legible!

```{r}
# YOUR CODE
# Would that I could. This is too difficult.
```

And to submit this rmarkdown, knit it into html. But first, clean up the code chunks, adjust the date, rename the author and change the `eval=FALSE` flag to `eval=TRUE` so your script actually generates an output.
Well done!


