---
title: "Make Countries Move"
author: "Laura Emilie Hollænder Jensen"
date: "20/3/2026"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#  Explore global development with R
In this exercise, you will load a filtered `gapminder` dataset -  with a subset of data on global development from 1952 - 2007 in increments of 5 years - to capture the period between the Second World War and the Global Financial Crisis. 

**Your task: Explore the data and visualise it in both static and animated ways, providing answers and solutions to 7 questions/tasks within this script.** 

- **Submit to Github this script with solutions and its knitted version.**  
- **Submit to Brightspace a document with the (1) URL of this script and its knitted version in Github (2) answer of 250 words and a (3) visualisation of the Final Question**

## Get the necessary packages
First, start with installing and activating the relevant packages `tidyverse`, `gganimate`, and `gapminder` if you do not have them already. 
*Pay attention to what warning messages you get when installing `gganimate`, as your computer might need other packages than `gifski` and `av`*

```{r libraries, echo = FALSE}
# install.packages("gganimate")
# install.packages("gifski")
# install.packages("av")
# install.packages("gapminder")
library(tidyverse)
library(gganimate)
library(gifski)
library(av)
library(gapminder)
```

## Look at the data and tackle the tasks
First, see which specific years are actually represented in the dataset and what variables are being recorded for each country. Note that when you run the cell below, Rmarkdown will give you two results - one for each line - that you can flip between.
```{r}
str(gapminder)
unique(gapminder$year)
head(gapminder)
glimpse(gapminder)
```

The dataset contains information on each country in the sampled year, its continent, life expectancy, population, and GDP per capita.

Let's plot all the countries in 1952.
```{r 1957}
theme_set(theme_bw())  # set theme to white background for better visibility

options(scipen = 999)

ggplot(subset(gapminder, year == 1952), aes(gdpPercap, lifeExp, size = pop, color = continent)) +
  geom_point() +
  scale_size(labels = scales::comma) +
  scale_x_log10(labels = scales::comma, 
  breaks = c(500, 1000, 2000, 5000, 10000, 20000, 50000)
  ) +
  ggtitle("Figure 01") +
  labs(
    title = "Life expectancy and wealth 1952",
    x = "GDP per Capita (log scale) in USD",
    y = "Life Expectancy in years",
    size = "Population",
    color = "Continent"
  ) +
    guides(color = guide_legend(override.aes = list(size = 6))) # bigger continent labels without bigger bubbles in the plot

gapminder %>% 
  filter(year == 1952) %>%
  arrange(desc(gdpPercap)) %>%
  head(5)
  
```

...

We see an interesting spread with an outlier to the right. Explore who it is so you can answer question 2 below!


Next, you can generate a similar plot for 2007 and compare the differences
```{r 2007}
ggplot(subset(gapminder, year == 2007), aes(gdpPercap, lifeExp, size = pop, color = continent)) +
  geom_point() +
  scale_size(labels = scales::comma) +
  scale_x_log10(labels = scales::comma,
  breaks = c(500, 1000, 2000, 5000, 10000, 20000, 50000)
  ) +
  ggtitle("Figure 02") +
    labs(
    title = "Life expectancy and wealth 2007",
    x = "GDP per Capita (log scale) in USD",
    y = "Life Expectancy in years",
    size = "Population",
    color = "Continent"
  ) +
  guides(color = guide_legend(override.aes = list(size = 6))) # bigger continent labels without bigger bubbles in the plot


gapminder %>% 
  filter(year == 2007) %>% 
  arrange(desc(gdpPercap)) %>% 
  head(5)
```

...

The black bubbles are a bit hard to read, the comparison would be easier with a bit more visual differentiation.

### Questions for the static figures:

1. *Answer: why does it make sense to have a log10 scale (`scale_x_log10()`) on the x axis?* (hint: try to comment it out and observe the result)

**LAURA'S ANSWER**: If we didn't use the logarithmic scale, almost all countries would cluster together far to the left, because Kuwait is such a big outlier that any differences in wealth between the other countries appear miniscule. Using the logarithmic scale spreads out the data points and makes change over time more visible. 

2. *Answer: In Figure 1: Who is the outlier (the richest country in 1952) far right on the x axis?* 

**LAURA'S ANSWER**: The outlier (the richest country in 1952) is Kuwait. We can find this out by taking a look at the gapminder dataset, using 'filter()' to show only rows that say '1952' in the year-column, and then sorting the data by GDP per capita in a descending order.

3. *Fix Figures 1 and 2: Differentiate the **continents** by color, and fix the axis labels and units to be more legible (**Hint**: the 2.50e+08 is so called "scientific notation". You want to eliminate it.)*

**LAURA'S ANSWER**: Taking directions from Microsoft Copilot, I used the command 'options(scipen = 999)' to tell R to avoid scientific notation, but I feel sketchy about it, because conceptually I don't really understand the command. 

4. *Answer: What are the five richest countries in the world in 2007?*

**LAURA'S ANSWER**: Using the same functions as I used to answer question 2, I can find the five richest countries in 2007 by filtering to show only data from 2007 and then sort it in a descending order by GDP per capita. Finally, running 'head(5)' shows me that the five countries with the highest GDP per capita in 2007 were Norway, Kuwait, Singapore, the US and Ireland.


##  Make it move!

The comparison would be easier if we had the two graphs together, animated. We have a lovely tool in R to do this: the `gganimate` package. Beware that there may be other packages your operating system needs in order to glue interim images into an animation or video. Read the messages when installing the package.

Also, there are *two* ways of animating the gapminder ggplot.

### Option 1: Animate using transition_states() 

The first step is to create the object-to-be-animated
```{r anim1 start}
anim <- ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop, color = continent)) +
  geom_point() +
  scale_size(labels = scales::comma) +
  scale_x_log10(labels = scales::comma, # convert to log-scale and add commas to make large numbers more legible
  breaks = c(500, 1000, 2000, 5000, 10000, 20000, 50000) # adding manual breaks to show more meaningful values on x-axis
  )  
  labs(
    title = "Year: {closest_state}",
    x = "GDP per Capita (log scale) in USD",
    y = "Life Expectancy in years"
  ) +
  guides(color = guide_legend(override.aes = list(size = 6))) # bigger continent labels without bigger bubbles in the plot

anim
```

...

This plot collates all the points across time. The next step is to split it into years and animate it. This may take some time, depending on the processing power of your computer (and other things you are asking it to do). Beware that the animation might appear in the bottom right 'Viewer' pane, not in this rmd preview. You need to `knit` the document to get the visual inside an html file.

```{r anim1}
anim + transition_states(year, 
                      transition_length = 1,
                      state_length = 1)

```
...

Notice how the animation moves jerkily, 'jumping' from one year to the next 12 times in total. This is a bit clunky, which is why it's good we have another option. 


### Option 2 Animate using transition_time()
This option smooths the transition between different 'frames', because it interpolates and adds transitional years where there are gaps in the timeseries data.

```{r anim2}
anim2 <- ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop, color = continent)) +
  geom_point() +
  scale_size(labels = scales::comma) +
  scale_x_log10(labels = scales::comma, # convert to log-scale, add commas to make the large numbers more legible
    breaks = c(500, 1000, 2000, 5000, 10000, 20000, 50000)  # adding manual breaks for more meaningful values on x-axis 
  ) + 
  transition_time(year) +
  labs(title = "Year: {frame_time}",
      x = "GDP per Capita (log scale) in USD",
      y = "Life Expectancy in years") +
  guides(color = guide_legend(override.aes = list(size = 6))) # bigger continent labels without bigger bubbles in the plot

anim2
```

The much smoother movement in Option 2 will be much more noticeable if you add a title to the chart, that will page through the years corresponding to each frame.

Now, choose one of the animation options and get it to work. You may need to troubleshoot your installation of `gganimate` and other packages 

### Tasks for the animations:

5. *Can you add a title to one or both of the animations above that will change* 
 *in sync with the animation?* *(**Hint**: search labeling for `transition_states()` and `transition_time()` functions respectively)*

**LAURA'S ANSWER**: I think I figured this one out, but once again I have to stress I used Copilot excessively and would not confidently be able to reproduce this code if stranded in the wilderness with no WiFi. I have added the necessary commands in the code chunks above. 


6. *Can you made the axes' labels and units more readable? Consider expanding the abbreviated labels as well as the scientific notation in the legend and x axis to whole numbers. Also, differentiate the countries from different continents by color*

**LAURA'S ANSWER**: I already got rid of the scientific notation, and adding 'color = continent' to the 'aes()' part of the code will colour the countries by continent. 


### Final Question
7. **Is the world a better place today than it was in the year you were born?** *Answer this question using the gapminder data.  Define better either as more prosperous, more free, more healthy, or suggest another measure that you can get from gapminder. Submit a 250 word answer with an illustration to Brightspace. Include a URL in your Brightspace submission that links to the coded solutions in Github. [Hint: if you wish to have more data than is in the filtered gapminder, you can load either the `gapminder_unfiltered` dataset or download more historical data at https://www.gapminder.org/data/ ]*

There is no data from my birth year (2000), so I am opting for the one that's closest, which I can find using first 'distinct(year)' to avoid seeing repetitions of the same year and 'arrange(desc())' to sort the years in descending order. Running these functions on the data shows me that 1997 and 2002 are the two years closest to 2000 where wealth and life expectancy was measured.
```{r years}
gapminder %>% 
  distinct(year) %>% 
  arrange(desc(year))

gapminder %>% 
  arrange(desc(pop))


```

I will choose the year 2002 as a proxy for my birth year, because it's the closest. Here, I am making a plot just like the ones for 1952 and 2007, but this time I will adjust the font size and size of the points to make the plot more legible in a png.  
```{r 2002}
theme_set(theme_bw())

ggplot(subset(gapminder, year == 2002), aes(gdpPercap, lifeExp, size = pop, color = continent)) +
  geom_point() +
  scale_size(labels = scales::comma) + # adding commas to make the large numbers more legible
  scale_x_log10(labels = scales::comma, # adding commas to make the large numbers more legible +
  breaks = c(500, 1000, 2000, 5000, 10000, 20000, 50000)  # adding manual breaks for more meaningful values on x-axis 
  ) + 
  ggtitle("Figure 02") +
    labs(
    title = "Life expectancy and wealth 2002",
    x = "GDP per Capita (log scale) in USD",
    y = "Life Expectancy in years",
    size = "Population", # renaming the size guide 'Population' instead of keeping the column name 'pop'
    color = "Continent" 
  ) +
    guides(
    color = guide_legend(override.aes = list(size = 8))  # bigger continent bubbles without bigger pop bottles
  ) 

ggsave("gapminder2002.png", width = 15, height = 10, dpi = 300) # saving image of plot
```

Another option is to use the 'gapminder_unfiltered'-data to see data from 2000. This data is more sparse and will not show me as much. 
```{r 2000}
theme_set(theme_bw())

ggplot(subset(gapminder_unfiltered, year == 2000), aes(gdpPercap, lifeExp, size = pop, color = continent)) +
  geom_point() +
  scale_size(labels = scales::comma) + # adding commas to make the large numbers more legible
  scale_x_log10(labels = scales::comma,# adding commas to make the large numbers more legible
  breaks = c(500, 1000, 2000, 5000, 10000, 20000, 50000) # adding manual breaks for more meaningful values on x-axis 
  ) + 
  ggtitle("Figure 02") +
    labs(
    title = "Life expectancy and wealth 2000",
    x = "GDP per Capita (log scale) in USD",
    y = "Life Expectancy in years",
    size = "Population", # renaming the size guide 'Population' instead of keeping the column name 'pop'
    color = "Continent" 
  ) +
    guides(
    color = guide_legend(override.aes = list(size = 6))  # bigger continent bubbles without bigger pop bottles
  )

ggsave("gapminder2000.png", width = 15, height = 10, dpi = 300) # saving image of plot
```

