---
title: "Week 13 Assignment -- Tell a story through a map"
author: "Laura Emilie Hollænder Jensen"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
date: "2026-03-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Assignment description
Following the Leaflet tutorial, develop a mini-project where you harvest or create digital data of your own, and plot these in an interactive map.
If using Rmarkdown, please post both the .rmd and the rendered .html files in your au##### repository. Submit here a *link* to your au##### repository in https://github.com/Digital-Methods-HASS which leads directly to the place where you have posted your .Rmd and .html files with the solutions.

```{r libraries, echo = FALSE}
library(leaflet)
library(tidyverse)
library(readr)
library(htmltools)
library(htmlwidgets) 
library(googlesheets4)
```

## Introduction to Laura's solution
For this assignment, I have decided to map the Danish market towns, or "købstæder". The market towns were towns which enjoyed special, royally issued privileges, namely a monopoly on craft and commerce. This meant that any journeyman wanting to practice his trade (e.g. a baker, a shoemaker, a milliner, etc.) had to set up shop in a market town, and any farmer wanting to sell his harvest had to travel to a market town to do so (and pay a toll to get into the town). Most of the market town privileges were repealed with the passing of "Næringsfrihedsloven" in 1857, but they remained as administrative areas until 1970 (Lex.dk). The last town to gain the status of købstad was Skjern in 1958.

### Task 1: Create a Danish equivalent of AUSmap with Esri layers, but call it DANmap.
```{r task 1, echo=TRUE}

DANmap <- leaflet() %>%   # assign the base location to an object
  setView( 10.85089,55.2339084, zoom = 13) %>% 
  addTiles()
DANmap

```

###  Task 2: Read in the googlesheet data into your DANmap object (with multiple background layers you created in Task 1).
```{r task 2, echo=TRUE}

# Reading in the Google sheet with the list of market towns.
gs4_deauth()  

market_towns <- read_sheet(
  "https://docs.google.com/spreadsheets/d/1t9yVSMgYTWAq4ql0xKvYuS-mJBSPCTRwhzY-VVm0LX8/edit?usp=sharing",
  sheet = "Ark2")

glimpse(market_towns)

# I need to change the Latitude and Longitude columns to numeric. 
market_towns$Latitude  <- as.numeric(market_towns$Latitude)
market_towns$Longitude <- as.numeric(market_towns$Longitude)
market_towns$Year_privileges_issued <- as.numeric(market_towns$Year_privileges_issued)


# Using addMarkers() to load the coordinates into a basic map.
market_town_map <- leaflet() %>% 
  addTiles() %>% 
  addMarkers(lng = market_towns$Longitude, 
             lat = market_towns$Latitude,
             popup = paste0("<b>", market_towns$By, "</b><br>",
                            ifelse(
                              is.na(market_towns$Year_privileges_issued),
                              "Year of købstad status unknown",
                              paste0("Got status as købstad in ", 
                                     market_towns$Year_privileges_issued))
                            ))
saveWidget(market_town_map, "market_town_map.html", selfcontained = T)

DANmap %>% 
  addMarkers(lng = market_towns$Longitude, 
             lat = market_towns$Latitude,
             popup = paste0("<b>", market_towns$By, "</b><br>",
                            ifelse(
                              is.na(market_towns$Year_privileges_issued),
                              "Year of købstad status unknown",
                              paste0("Got status as købstad in ", 
                                     market_towns$Year_privileges_issued))
                            ))

```


### Task 3: Can you cluster the points in Leaflet?
```{r task 3, echo=TRUE}

DANmap %>% 
  addMarkers(lng = market_towns$Longitude, 
             lat = market_towns$Latitude,
             popup = paste0("<b>", market_towns$By, "</b><br>",
                            ifelse(
                              is.na(market_towns$Year_privileges_issued),
                              "Year of købstad status unknown",
                              paste0("Got status as købstad in ", 
                                     market_towns$Year_privileges_issued))
                            ), 
             clusterOptions = markerClusterOptions())

```


### Task 4: Look at the two maps (with and without clustering) and consider what each is good for and what not.
At first glance, the clustered map offers a more clear and less messy view of the Danish market towns, since they are scattered all over the country and make the map look crowded. However, the clustered map is not perfect, because it risks being misleading. For examle, if I zoom out and look at Fyn, I will see a green cluster with the number '17'. Instinctively, that makes me think there must be 17 market towns on Fyn. But, if I zoom in enough, I can see that this is not so: There are indeed 17 market towns in the cluster, but some of them are in Jylland. In short: The clustered map is imperfect because it fails to take into account that a Danish person's prior knowledge and expectations will probably cause them to 'group' a Danish map into three parts: Jylland, Fyn and Sjælland.

### Task 5: Find out how to display the notes and classifications column in the map. 
I have added a column containing the (ESTIMATED) year each town gained status as "market town". These have already been displayed through the commands I ran in task 2 with the added explanatory note: "Got status as købstad in" to make the popup more clear. 

